---
title: "Metadata"
date: "`r Sys.Date()`"
---

```{r knitropts, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy = "styler",
  warning = FALSE,
  message = FALSE
)

library(climaemet)
library(dplyr)
```

Metadata of **climaemet** functions as of **`r format(Sys.Date(),"%d %b %Y")`**:

## `aemet_daily`

See `?aemet_daily` help page.

<details>

<summary>Show metadata</summary>

```{r , echo=FALSE, results='asis'}
meta <- aemet_daily_clim(extract_metadata = TRUE)
knitr::kable(meta$campos)
```

</details>

## `aemet_extremes_clim`

See `?aemet_extremes_clim` help page.

<details>

<summary>Show metadata</summary>

```{r , echo=FALSE, results='asis'}
meta1 <- aemet_extremes_clim("9434", "V", extract_metadata = TRUE)$campos
meta2 <- aemet_extremes_clim("9434", "T", extract_metadata = TRUE)$campos
meta3 <- aemet_extremes_clim("9434", "P", extract_metadata = TRUE)$campos

meta1 %>%
  mutate(param = "V") %>%
  relocate(param, .before = 1) %>%
  bind_rows(meta2 %>% mutate(param = "T")) %>%
  bind_rows(meta3 %>% mutate(param = "P")) %>%
  knitr::kable()
```

</details>

## `aemet_forecast_daily`

See `?aemet_forecast_daily` help page.

<details>

<summary>Show metadata</summary>

```{r , echo=FALSE, results='asis'}
meta <- aemet_forecast_daily("08001", extract_metadata = TRUE)
knitr::kable(meta)
```

</details>

## `aemet_forecast_hourly`

See `?aemet_forecast_hourly` help page.

<details>

<summary>Show metadata</summary>

```{r , echo=FALSE, results='asis'}
meta <- aemet_forecast_hourly("08001", extract_metadata = TRUE)
knitr::kable(meta)
```

</details>

## `aemet_lastobs`

See `?aemet_last_obs` help page.

<details>

<summary>Show metadata</summary>

```{r , echo=FALSE, results='asis'}
meta <- aemet_last_obs(extract_metadata = TRUE)
knitr::kable(meta$campos)
```

</details>

## `aemet_monthly`

See `?aemet_monthly` help page.

<details>

<summary>Show metadata</summary>

```{r , echo=FALSE, results='asis'}
tday <- Sys.Date()
year <- format(tday, "%Y")

apidest <- paste0(
  "/api/valores/climatologicos/mensualesanuales/datos/anioini/",
  year, "/aniofin/", year, "/estacion/9434"
)

meta <- get_metadata_aemet(apidest)
knitr::kable(meta$campos)
```

</details>

## `aemet_normal`

See `?aemet_normal` help page.

<details>

<summary>Show metadata</summary>

```{r , echo=FALSE, results='asis'}
apidest <- paste0("/api/valores/climatologicos/normales", "/estacion/9434")

meta <- get_metadata_aemet(apidest)
knitr::kable(meta$campos)
```

</details>
